{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-19 thesis/overview.tex
daves final revisions to overview, as well as some of myown tweaks

\subsubsection{Shared state}
When a programming language offers multithreaded parallelism but forbids access to any shared state between threads \cite{rust-book},
the simultaneity of threads is largely irrelevant to the program's behaviour.
However, ``thread-unsafe'' languages such as C, C++, Java, and so on remain popular,
in which threads may access global or heap-allocated variables and data structures with no enforced access discipline.
The behaviour of such programs is then subject to the manner in which these accesses interleave.

@@

\begin{figure}[t]
	\begin{tabular}{cc}
		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
			{\small
			\begin{tabular}{l}
				\texttt{\ctype{int} x;} \\
				\texttt{\ctype{void} \call{count}() \{} \\
				\texttt{~~~~\call{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++)} \\
				\texttt{~~~~~~~~x++;} \\
				\texttt{\}} \\
				\texttt{\ctype{void} \call{main}() \{} \\
				\texttt{~~~~tid1 = \call{thr\_create}(count);} \\
				\texttt{~~~~tid2 = \call{thr\_create}(count);} \\
				\texttt{~~~~\call{thr\_join}(tid1);} \\
				\texttt{~~~~\call{thr\_join}(tid2);} \\
				\texttt{~~~~\call{printf}(\const{"\%d\textbackslash{}n"}, x);} \\
				\texttt{\}} \\
			\end{tabular}
			}
			&
			{\small \begin{tabular}{ll}
				{\bf Thread 1} & {\bf Thread 2} \\
				\hline
				\texttt{load tmp <- x;} & \\
				& \texttt{load tmp <- x;} \\
				& \texttt{add tmp <- 1;} \\
				& \texttt{store x <- tmp;} \\
				\texttt{add tmp <- 1;} & \\
				\texttt{store x <- tmp;} & \\
			\end{tabular}
			}
			\\
			\\
			(a) Source listing for a multithreaded program which might count to 2000.
			&
			% revisionminor can't go here i guess
			(b) Example interleaving of (a)'s compiled assembly,
			in which two concurrent iterations of the loop yield one net increment of {\tt x}.
		\end{tabular}
	\end{tabular}
	\caption[Example concurrent program.]{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
	\label{fig:concurrency-bug}
\end{figure}

@@

\begin{figure}[h]
        \small
	\begin{center}
\begin{tabular}{c}
\begin{tabular}{rll}
	& \multicolumn{2}{c}{\texttt{\ctype{int} x = \const{0}; \ctype{bool} y = \const{false}; \ctype{mutex\_t} mx;}} \\
        \\
        & {\bf Thread 1} & {\bf Thread 2} \\
	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// A1}} & \\
	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
	3 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
	4 & & \texttt{\call{mutex\_lock}(\&mx);} \\
	5 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
	6 & & \texttt{\hilight{brickred}{x++;}~\ccomment{// A2}} \\
\end{tabular}
\\
\\
	{\normalsize (a) True potential data race.}
\\
\\
\begin{tabular}{rll}
        %& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
        & {\bf Thread 1} & {\bf Thread 2} \\
	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// B1}} & \\
	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
	3 & \texttt{y = \const{true};} & \\
	4 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
	5 & & \texttt{\call{mutex\_lock}(\&mx);} \\
	6 & & \texttt{\ctype{bool} tmp = y;} \\
	7 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
	8 & & \texttt{\flow{if} (tmp) \hilight{brickred}{x++;}~\ccomment{// B2}} \\
        %8 & & \texttt{if (tmp)} \\
        %9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
\end{tabular}
\\
\\
{\normalsize (b) No data race in any interleaving.}
\end{tabular}
	\end{center}
	\caption[Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.]
		{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
	Applying Happens-Before to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
	while using Limited Happens-Before on (b) will produce a false alarm on B1/B2.}}
\label{fig:hb-example}
\end{figure}

@@

The Pebbles kernel architecture
is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and \revisionminor{Implementation} \cite{kspec,thrlib}.
In the course of a semester, students work on five programming assignments;
the first two are individual, and the remaining three are the products of two-person teams.
I will focus on the third and fourth of these, the thread library and kernel,
called ``P2'' and ``P3'' respectively (the project numbers start at 0).
The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
The course's prerequisite is 15-213, Introduction to Computer Systems \cite{csapp,sigcse01:CSaPP}.
Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in \Cref{tab:syscalls}.

@@

\subsubsection{P2}
The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
\revision{Students implement it on top of a reference Pebbles kernel implementation (named Pathos~\cite{pathos}),
which is distributed as an obfuscated binary to protect the subsequent challenges of P3.}
The required concurrency primitives \revisionminor{for P2} are as follows:
\begin{itemize}
	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter. Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal} {\tt (cvp)}, and {\tt cond\_broadcast(cvp)}. {\tt cond\_wait} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls to signal or broadcast under that mutex).
		{\tt cond\_signal} and {\tt cond\_broadcast} wake one or all waiting threads.
		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)}
		(sometimes called \revisionminor{$P$ and $V$ in other literature,
		from Dutch,
		originally {\em passering} and {\em vrijgave} \cite{ewd35},
		later often explained as {\em proberen} and {\em verhogen}}).
		The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
	\item Reader-writer locks (rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
		Students typically implement rwlocks using mutexes and condition variables, not using atomic instructions or system calls directly.
\end{itemize}
The interface to each also includes an associated {\tt \_init()} and {\tt \_destory()} function.

@@

\subsubsection{P3}
In P3, students implement a kernel which provides the same system calls shown in \Cref{tab:syscalls}, previously provided by the reference kernel.
Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
This requires less implementation complexity than the more featureful
\revisionminor{{\tt rfork} model, of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model, of Linux}.

@@

		The set of function signatures which comprise the interface to a body of code.

@@

		Also, this thesis is focused on identifying failures, not root causes,

@@

		where each thread is blocked waiting for an event that can be caused only by another thread in the set

@@

		Thread library for Linux programs provided by POSIX.

@@

		Mach's C threads \cite{mach-c-threads}
		and DEC's Systems Research Center threads \cite{src-threads},

>>>>

\subsubsection{Shared state}
When a programming language offers multithreaded parallelism but forbids access to any shared state between threads
\cite{erlang,rust-book},
the simultaneity of threads is largely irrelevant to the program's behaviour.
However, ``thread-unsafe'' languages such as C, C++, Java, and so on remain popular,
in which threads may access global or heap-allocated variables and data structures with no enforced access discipline.
The behaviour of such programs is then subject to the manner in which these accesses interleave.

@@

\begin{figure}[t]
	\begin{tabular}{cc}
		\begin{tabular}{p{0.45\textwidth}p{0.5\textwidth}}
			{\small
			\begin{tabular}{l}
				\texttt{\ctype{int} x = \const{0};} \\
				\texttt{\ctype{void} \call{count}() \{} \\
				\texttt{~~~~\call{for} (\ctype{int} i = \const{0}; i < \const{1000}; i++)} \\
				\texttt{~~~~~~~~x++;} \\
				\texttt{\}} \\
				\texttt{\ctype{void} \call{main}() \{} \\
				\texttt{~~~~tid1 = \call{thr\_create}(count);} \\
				\texttt{~~~~tid2 = \call{thr\_create}(count);} \\
				\texttt{~~~~\call{thr\_join}(tid1);} \\
				\texttt{~~~~\call{thr\_join}(tid2);} \\
				\texttt{~~~~\call{printf}(\const{"\%d\textbackslash{}n"}, x);} \\
				\texttt{\}} \\
			\end{tabular}
			}
			&
			{\small \begin{tabular}{ll}
				{\bf Thread 1} & {\bf Thread 2} \\
				\hline
				\texttt{load tmp <- x;} & \\
				& \texttt{load tmp <- x;} \\
				& \texttt{add tmp <- 1;} \\
				& \texttt{store x <- tmp;} \\
				\texttt{add tmp <- 1;} & \\
				\texttt{store x <- tmp;} & \\
			\end{tabular}
			}
			\\
			\\
			(a) Source listing for a multithreaded program which might count to 2000.
			&
			% revisionminor can't go here i guess
			(b) Example interleaving of (a)'s compiled assembly,
			in which two concurrent iterations of the loop yield one net increment of {\tt x}.
		\end{tabular}
	\end{tabular}
	\caption[Example concurrent program.]{Example concurrent program in which simultaneous accesses to shared state may interleave to produce unexpected results.}
	\label{fig:concurrency-bug}
\end{figure}

@@

\begin{figure}[h]
        \small
	\begin{center}
\begin{tabular}{c}
\begin{tabular}{rll}
	\\
	& \multicolumn{2}{c}{\texttt{\ctype{int} x = \const{0}; \ctype{bool} y = \const{false}; \ctype{mutex\_t} mx;}} \\
        \\
        & {\bf Thread 1} & {\bf Thread 2} \\
	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// A1}} & \\
	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
	3 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
	4 & & \texttt{\call{mutex\_lock}(\&mx);} \\
	5 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
	6 & & \texttt{\hilight{brickred}{x++;}~\ccomment{// A2}} \\
\end{tabular}
\\
\\
	{\normalsize (a) True potential data race.}
\\
\\
\begin{tabular}{rll}
        %& \multicolumn{2}{c}{\texttt{int x = 0; bool y = false; mutex\_t mx;}} \\
        & {\bf Thread 1} & {\bf Thread 2} \\
	1 & \texttt{\hilight{brickred}{x++;}~\ccomment{// B1}} & \\
	2 & \texttt{\call{mutex\_lock}(\&mx);} & \\
	3 & \texttt{y = \const{true};} & \\
	4 & \texttt{\call{mutex\_unlock}(\&mx);} & \\
	5 & & \texttt{\call{mutex\_lock}(\&mx);} \\
	6 & & \texttt{\ctype{bool} tmp = y;} \\
	7 & & \texttt{\call{mutex\_unlock}(\&mx);} \\
	8 & & \texttt{\flow{if} (tmp) \hilight{brickred}{x++;}~\ccomment{// B2}} \\
        %8 & & \texttt{if (tmp)} \\
        %9 & & \texttt{~~~~\hilight{brickred}{x++;}~// B2} \\
\end{tabular}
\\
\\
{\normalsize (b) No data race in any interleaving.}
\end{tabular}
	\end{center}
	\caption[Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.]
		{{Data-race analyses may be prone to either {\em false negatives} or {\em false positives}.
	Applying Happens-Before to program (a) will miss the potential race possible between A1/A2 in an alternate interleaving,
	while using Limited Happens-Before on (b) will produce a false alarm on B1/B2.}}
\label{fig:hb-example}
\end{figure}

@@

The Pebbles kernel architecture
is used at Carnegie Mellon University (CMU) in 15-410, Operating System Design and \revisionminor{Implementation} \cite{kspec,thrlib}.
In the course of a semester, students work on five programming assignments;
the first two are individual, and the remaining three are the products of two-person teams.
I will focus on the third and fourth of these, the thread library and kernel,
called ``P2'' and ``P3'' respectively (the project numbers start at 0).
The other three (a stack-crawling backtrace utility, a bare-metal game with device drivers, and a small extension to the P3 kernel) are not of concern in this thesis.
The course's prerequisite is 15-213, Introduction to Computer Systems \cite{sigcse01:CSaPP,csapp}.
Both P2 and P3 are built using the {\em Pebbles} system call specification, outlined in \Cref{tab:syscalls}.

@@

\subsubsection{P2}
The thread library project \cite{thrlib} has two main components: implementing concurrency primitives, and implementing thread lifecycle and management routines.
\revision{Students implement it on top of a reference Pebbles kernel implementation (named Pathos~\cite{pathos}),
which is distributed as an obfuscated binary to protect the subsequent challenges of P3.}
The required concurrency primitives \revisionminor{for P2} are as follows:
\begin{itemize}
	\item Mutexes, with the interface {\tt mutex\_lock(mp)} and {\tt mutex\_unlock(mp)}, whose functionality is described earlier this chapter. Students may use any x86 atomic instruction(s) they desire, such as {\tt xchg}, {\tt xadd}, or {\tt cmpxchg}, and/or the {\tt deschedule}/ {\tt make\_runnable} system calls offered by the reference kernel.
	\item Condition variables, with the interface {\tt cond\_wait(cvp, mp)}, {\tt cond\_signal(cvp)}, and {\tt cond\_broadcast(cvp)}. {\tt cond\_wait()} blocks the invoking thread, ``simultaneously'' releasing a mutex which protects some associated state (atomically, with respect to other calls under the same mutex).
		{\tt cond\_signal()} and {\tt cond\_broadcast()} wake one or all waiting threads.
		Students must use the {\tt deschedule} and {\tt make\_runnable} system calls to implement blocking (busy-waiting is forbidden), and typically include an internal mutex to protect the condition variable's state as well.
		The primary challenge of this exercise is ensuring the aforementioned atomicity between {\tt cond\_wait}'s unlock and deschedule, with respect to the rest of the interface.
	\item Semaphores, with the interface {\tt sem\_wait(sp)} and {\tt sem\_signal(sp)}
		(sometimes called \revisionminor{$P$ and $V$ in other literature,
		from Dutch,
		originally {\em passering} and {\em vrijgave} \cite{ewd35},
		later often explained as {\em proberen} and {\em verhogen}}).
		The semaphore can be initialized to any integer value; if initialized to 1, it behaves like a mutex.
		Students typically implement semaphores using mutexes and condition variables, not using atomic instructions or system calls directly.
	\item Reader-writer locks (also R/W-locks or rwlocks), with the interface {\tt rwlock\_lock(rwp, mode)} and {\tt rwlock\_unlock(rwp)}. {\tt mode} may be either {\tt RWLOCK\_READ} or {\tt RWLOCK\_\allowbreak{}WRITE}.
		Behaves as mutexes, but multiple readers may access the critical section simultaneously.
		Students typically implement rwlocks using mutexes and condition variables, not using atomic instructions or system calls directly.
\end{itemize}
The interface to each also includes an associated {\tt \_init()} and {\tt \_destory()} function.

@@

\subsubsection{P3}
In P3, students implement a kernel which provides the same system calls shown in \Cref{tab:syscalls}, previously provided by the reference kernel.
Pebbles adopts the Mach \cite{DBLP:conf/usenix/AccettaBBGRTY86} distinction between {\em tasks}, which are resource containers, and {\em threads}, each of which executes within a single task.
This requires less implementation complexity than the more featureful
\revisionminor{{\tt rfork} model of Plan 9 \cite{Pike90plan9}, or the {\tt clone} model of Linux}.

@@

		A set of function signatures which comprise the interface to a body of code.

@@

		Also, this thesis is focused on identifying failures,
		leaving root cause diagnosis to the user,
		%not root causes,

@@

		where each is blocked waiting for an event that can be caused only by another thread in the set

@@

		POSIX thread library for Linux programs. %provided by POSIX.

@@

		DEC's Systems Research Center threads \cite{src-threads}
		and Mach's C threads \cite{mach-c-threads},
